rOpenSci tools for accessing research literature for text mining

### 2019-04-08 manuscript.Rmd
paper updates

---
layout: review, 11pt
linenumbers: true
title: "rOpenSci tools for accessing science literature for textmining"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

@@

@@

### Case study 1: Citation mining

@@

### Case study 2: Abstract mining

@@

### Case study 3: Full text mining

>>>>

---
layout: review, 11pt
linenumbers: true
title: "rOpenSci tools for accessing research literature for text mining"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

@@

# fulltext: a swiss army knife for text mining in R

[fulltext][] is a general purpose R package for the data part of text-mining:
search for articles, get links to articles, get article abstracts, and fetch
full text of articles. The `fulltext` package is always adding additional
data sources as time allows (See Table 1). Starting from searching for 
articles, the outputs of search can be fed into a function to get links 
to those articles, or to get abstracts for those articles, or to fetch 
their full text.

The following is a breakdown of the major distinct functional parts of 
`fulltext`.

## Search

`ft_search()` provides search access to nine different data sources (PLOS, BMC, Crossref,
Entrez, arXiv, bioRxiv, Europe PMC, Scopus, Microsoft Academic), creating a
mostly unified interface to all data sources. The parts of each data source
that are common are mostly factored into the parameters of the `ft_search()` function,
and we also allow the user to pass on data source specific options as needed.

## Links

`ft_links()` provides two pathways to get links (URLs) for articles, with a choice
of four different data sources (PLOS, BMC, Crossref, Entrez). First, you can
use `ft_search()`, then pass the output of that function to `ft_links()`. Second, 
you can pass DOIs directly to `ft_links()`. Both end up at the same point, links
for each article, if they could be found for the user selected data source.

The biggest caveat with `ft_links()` is that we can't gaurantee that the links
will work. Link rot is one way in which the links may not work: link rot is when 
the URL does not point to the original content anymore, or fails altogether. 
Additionally, with Crossref, publishers can deposit URLs for articles, but they 
make change the URLs at some later date but not update the URLs with Crossref.

## Abstracts

`ft_abstract()` provides access to article abstracts from four different 
data sources (PLOS, Scopus, Microsoft Academic Research, Crossref). The only 
way to use the function is to pass article identifiers, which are for the most 
DOIs.

The advantage of abstracts over full text is that abstracts can often be 
retrieved even for paywalled articles. That is, you can have much broader
coverage of the articles you're targeting relative to full text.

If you are after abstracts, and you are already getting or already have full text, 
and if the articles are in XML format, then you can use [pubchunks][] to extract
out the abstracts.

## Fetch full text

`ft_get()` fetchs full text of articles from many different data sources. From the DOIs
that are passed in to the function, we detect the publisher, and there are specific plugins
for certain publishers:

- aaas
- aip
- amersocclinoncol
- amersocmicrobiol
- arxiv
- biorxiv
- bmc
- copernicus
- crossref
- elife
- elsevier
- entrez
- frontiersin
- ieee
- informa
- instinvestfil
- jama
- microbiology
- peerj
- pensoft
- plos
- pnas
- royalsocchem
- sciencedirect
- scientificsocieties
- wiley

If there's no built-in plugin for the publisher already, we use the FTDOI API 
(<https://ftdoi.org>) to try to get the link for the full text of the article. If 
the FTDOI API doesn't bear fruit, we search Crossref for a link to the full text.
If Crossref doesn't have any full text links, we give up.

Since users can go through a lot of article requests, we cache successfully 
downloaded articles, and keep that knowledge consistent across R sessions; all 
subsequent requests for the same article just use the cached version. Additionally,
all errors in `ft_get()` are collected in a tidy data.frame in the output of the 
function to help the user quickly determine what went wrong.

@@

## Case study 1: Citation mining

@@

## Case study 2: Abstract mining

@@

## Case study 3: Full text mining

### 2019-04-08 manuscript.tex
paper updates

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={rOpenSci tools for accessing science literature for textmining},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{rOpenSci tools for accessing science literature for textmining}
    \author[cstar]{Scott Chamberlain\corref{c1}}
   \ead{myrmecocystus(at)gmail.com} 
   \cortext[c1]{Corresponding author}
      \address[cstar]{rOpenSci, Museum of Paleontology, University of California, Berkeley,
CA, USA}

@@

@@

\hypertarget{case-study-1-citation-mining}{%
\subsubsection{Case study 1: Citation
mining}\label{case-study-1-citation-mining}}

@@

\hypertarget{case-study-2-abstract-mining}{%
\subsubsection{Case study 2: Abstract
mining}\label{case-study-2-abstract-mining}}

@@

\hypertarget{case-study-3-full-text-mining}{%
\subsubsection{Case study 3: Full text
mining}\label{case-study-3-full-text-mining}}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={rOpenSci tools for accessing research literature for text mining},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{rOpenSci tools for accessing research literature for text mining}
    \author[cstar]{Scott Chamberlain\corref{c1}}
   \ead{myrmecocystus(at)gmail.com} 
   \cortext[c1]{Corresponding author}
      \address[cstar]{rOpenSci, Museum of Paleontology, University of California, Berkeley,
CA, USA}

@@

\hypertarget{fulltext-a-swiss-army-knife-for-text-mining-in-r}{%
\section{fulltext: a swiss army knife for text mining in
R}\label{fulltext-a-swiss-army-knife-for-text-mining-in-r}}

\href{https://github.com/ropensci/fulltext}{fulltext} is a general
purpose R package for the data part of text-mining: search for articles,
get links to articles, get article abstracts, and fetch full text of
articles. The \texttt{fulltext} package is always adding additional data
sources as time allows (See Table 1). Starting from searching for
articles, the outputs of search can be fed into a function to get links
to those articles, or to get abstracts for those articles, or to fetch
their full text.

The following is a breakdown of the major distinct functional parts of
\texttt{fulltext}.

\hypertarget{search}{%
\subsection{Search}\label{search}}

\texttt{ft\_search()} provides search access to nine different data
sources (PLOS, BMC, Crossref, Entrez, arXiv, bioRxiv, Europe PMC,
Scopus, Microsoft Academic), creating a mostly unified interface to all
data sources. The parts of each data source that are common are mostly
factored into the parameters of the \texttt{ft\_search()} function, and
we also allow the user to pass on data source specific options as
needed.

\hypertarget{links}{%
\subsection{Links}\label{links}}

\texttt{ft\_links()} provides two pathways to get links (URLs) for
articles, with a choice of four different data sources (PLOS, BMC,
Crossref, Entrez). First, you can use \texttt{ft\_search()}, then pass
the output of that function to \texttt{ft\_links()}. Second, you can
pass DOIs directly to \texttt{ft\_links()}. Both end up at the same
point, links for each article, if they could be found for the user
selected data source.

The biggest caveat with \texttt{ft\_links()} is that we can't gaurantee
that the links will work. Link rot is one way in which the links may not
work: link rot is when the URL does not point to the original content
anymore, or fails altogether. Additionally, with Crossref, publishers
can deposit URLs for articles, but they make change the URLs at some
later date but not update the URLs with Crossref.

\hypertarget{abstracts}{%
\subsection{Abstracts}\label{abstracts}}

\texttt{ft\_abstract()} provides access to article abstracts from four
different data sources (PLOS, Scopus, Microsoft Academic Research,
Crossref). The only way to use the function is to pass article
identifiers, which are for the most DOIs.

The advantage of abstracts over full text is that abstracts can often be
retrieved even for paywalled articles. That is, you can have much
broader coverage of the articles you're targeting relative to full text.

If you are after abstracts, and you are already getting or already have
full text, and if the articles are in XML format, then you can use
\href{https://github.com/ropensci/pubchunks}{pubchunks} to extract out
the abstracts.

\hypertarget{fetch-full-text}{%
\subsection{Fetch full text}\label{fetch-full-text}}

\texttt{ft\_get()} fetchs full text of articles from many different data
sources. From the DOIs that are passed in to the function, we detect the
publisher, and there are specific plugins for certain publishers:

\begin{itemize}
\tightlist
\item
  aaas
\item
  aip
\item
  amersocclinoncol
\item
  amersocmicrobiol
\item
  arxiv
\item
  biorxiv
\item
  bmc
\item
  copernicus
\item
  crossref
\item
  elife
\item
  elsevier
\item
  entrez
\item
  frontiersin
\item
  ieee
\item
  informa
\item
  instinvestfil
\item
  jama
\item
  microbiology
\item
  peerj
\item
  pensoft
\item
  plos
\item
  pnas
\item
  royalsocchem
\item
  sciencedirect
\item
  scientificsocieties
\item
  wiley
\end{itemize}

If there's no built-in plugin for the publisher already, we use the
FTDOI API (\url{https://ftdoi.org}) to try to get the link for the full
text of the article. If the FTDOI API doesn't bear fruit, we search
Crossref for a link to the full text. If Crossref doesn't have any full
text links, we give up.

Since users can go through a lot of article requests, we cache
successfully downloaded articles, and keep that knowledge consistent
across R sessions; all subsequent requests for the same article just use
the cached version. Additionally, all errors in \texttt{ft\_get()} are
collected in a tidy data.frame in the output of the function to help the
user quickly determine what went wrong.

@@

\hypertarget{case-study-1-citation-mining}{%
\subsection{Case study 1: Citation
mining}\label{case-study-1-citation-mining}}

@@

\hypertarget{case-study-2-abstract-mining}{%
\subsection{Case study 2: Abstract
mining}\label{case-study-2-abstract-mining}}

@@

\hypertarget{case-study-3-full-text-mining}{%
\subsection{Case study 3: Full text
mining}\label{case-study-3-full-text-mining}}
